\bf Adaption of the Chumbley Score to matching of bullet striation marks

### 2018-02-26 doc/paper.Rmd
up to  line 226

## Motivation
Same source analyses are a major part of an Forensic Toolmark examiners job. In current practice firearm examiners  make these comparisons visually under a comparison microscope and make one of the following four conclusions: identification, inconclusive, elimination or unsuitable for examination \citet{afte-website-2}. These conclusions are made on the basis of "unique surface contours" of the two toolmarks being in "sufficient agreement" \citep{afte-toolmarks1998}. AFTE describes the term "sufficient agreement" as the possibility of another tool producing the markings under comparison, as practically impossible \citep{afte-toolmarks1998} \citep{afte-website}. \gk{Although, they go on to concede that the interpretation of the identification depends on the examiner and is  therefore subjective in nature.}

One of the main criticisms in the toolmark examination first raised by the National Research Council in 2009 \citep{NAS:2009} and later emphasized further by the President's Council of Advisors on Science and Technology \citep{pcast2016} are the amount of subjectivity in the assessment and the lack of error rates.
Using a statistical approach removes both subjectivity from the assessment and allows us to provide error rates for both false positive and false negative identifications.

Digitized images, e.g. profilometers or confocal microscopy allow us to measure the surface \citep{vorburger2016} and can be used to conduct statistical analysis on the markings. Multiple toolmarks have been studied in the literature:  \citet{manytoolmarks1, chumbley} have studied screwdriver marks digitized using a profilometer,  \citet{manytoolmarks2} have investigated screwdriver, tongue and groove pliers 3D marks that are generated using a confocal microscope,  while, slip-joint pliers have been investigated by \citet{afte-chumbley} using digitized marks generated by a surface profilemeter. \citet{manytoolmarks2} define a relative distance metric and use it as similarity measure between two toolmarks. \citet{manytoolmarks1} used an algorithm which compares many small segments in the markings between two toolmarks. A maximum pearson correlation coefficient is then used as the statistical index for comparison. The Chumbley scoring method, first introduced by \citet{chumbley}, uses a similar but much more extenstive framework which includes a non-parametric statistical test. This was later improved by Hadler and Morris \citep{hadler}. In this paper, we are investigating the applicability of the Chumbley scoring method by \citet{hadler} to assessing striation marks on bullets for same-source identification.

Striation marks on bullets are made by impurities in the barrel. As the bullet travels through the barrel, these imperfections leave "scratches" on the bullet surface. Typically, only striation marks in the land engraved areas (LEAs) are considered \citet{afte-article1992}. Bullet Lands are depressed areas between bullet grooves made by the rifling action of the barrel. Compared to toolmarks made by screwdrivers striation marks on bullets are typically much smaller, both in length and in width. Bullets also have a curved cross-sectional topography.

>>>>

## Motivation
Same source analyses are a major part of an Forensic Toolmark Examiner's job. In current practice examiners  make these comparisons visually under a comparison microscope and come to one of the following four conclusions: identification, inconclusive, elimination or unsuitable for examination \citet{afte-website-2}. These conclusions are made on the basis of "unique surface contours" of the two toolmarks being in "sufficient agreement" \citep{afte-toolmarks1998}. AFTE describes the term "sufficient agreement" as the possibility of another tool producing the markings under comparison, as practically impossible \citep{afte-toolmarks1998} \citep{afte-website}. 
%\gk{Although, they go on to concede that the interpretation of the identification depends on the examiner and is  therefore subjective in nature.} 
This subjectivity in the assessment as well as the lack of error rates are the main points of criticisms in the toolmark examination first raised by the National Research Council in 2009 \citep{NAS:2009} and later emphasized further by the President's Council of Advisors on Science and Technology \citep{pcast2016}.

Technological advances, such as profilometers and confocal microscopy allow to measure 3D surfaces \citep{vorburger2016} in a digitized form. This forms the basis of statistical analysis of toolmarks. A statistical approach based on data  removes both subjectivity from the assessment and allows a quantification of error rates for both false positive and false negative identifications.

Various toolmarks have been studied in the literature:  screwdriver marks  digitized using a profilometer have been analyzed in \citet{manytoolmarks1} and \citet{chumbley};  \citet{manytoolmarks2} have investigated 3D marks from screwdriver, tongue and groove pliers captured using a confocal microscope;  digitized marks from slip-joint pliers generated by a surface profilometer have been investigated by \citet{afte-chumbley}.

\citet{manytoolmarks2} define a relative distance metric and use it as similarity measure between two toolmarks. \citet{manytoolmarks1} extract many small segments in the markings of two toolmarks and compare similiarity using a maximum pearson correlation coefficient. 
The Chumbley scoring method, first introduced by \citet{chumbley}, uses a similar but more extensive framework based on a Mann-Whitney U test of the resulting correlation coefficients. This approach was later improved by Hadler and Morris \citep{hadler}. \hh{XXX describe the improvement in half a sentence XXX} In this paper, we are investigating the applicability of the Chumbley scoring method by \citet{hadler} to assess striation marks on bullet lands for same-source identification.

Striation marks on bullets are made by impurities in the barrel. As the bullet travels through the barrel, these imperfections leave "scratches" on the bullet surface. Typically, only striation marks in the land engraved areas (LEAs) are considered \citet{afte-article1992}. Bullet lands are depressed areas between the  grooves made by the rifling action of the barrel. Compared to toolmarks made by screwdrivers striation marks on bullets are typically much smaller, both in length and in width. Bullets also have a curved cross-sectional topography \hh{sketch?}.

### 2018-02-26 doc/paper.tex
up to  line 226

\hh{XXX first sentence: Same source analyses are a major part of an Forensic Toolmark examiners job. XXX second sentence: describe standard practice: In current practice firearm examiners  make these comparisons visually under a comparison microscope and report XXX what is being reported: AFTE. XXX find: AFTE theory of identification as it relates to toolmarks  }
\hh{One of the main criticisms in this toolmark examination first raised by the National Research Council in 2009 [@NAS:2009] and later emphasized further by the President's Council of Advisors on Science and Technology [@pcast2016] are the amount of subjectivity in the assessment and the lack of error rates. }
\hh{XXX we have digitized images, e.g. profilometers or Confocal microscopy allow us to measure the surface}

@@

\hh{XXX In this pper, we are investigating the applicability of the Chumbley scoring method by @hadler to assessing striation marks on bullets for same-source identification. }

\hh{Striation marks on bullets are made by impurities in the barrel. As the bullet travels through the barrel, these imperfections leave "scratches' on the bullet surface. Typically, only striation marks in the land engraved areas (LEAs) are considered (XXX citation) [@afte-article1992]. Compared to toolmarks made by screwdrivers striation marks on bullets are typically much smaller, both in length and in width. Bullets also have a curved cross-sectional topography. 
}

Compairing pairs of toolmarks with the intention of matching it to a
tool has been studied many times in the past. Extensive examples can be
found in literature for tools and toolmark research ranging from
screwdrivers \citep{manytoolmarks1, manytoolmarks2, chumbley} to groove
pliers \citep{manytoolmarks2} to slip-joint pliers \citep{afte-chumbley}
and many more. In comparison to this, same source matching of bullets to
firearms has not been examined as prominently as that of toolmarks. Even
less information is available on validity of methods and error rates
associated with firearms examination. The National Academy of Sciences
in its report in 2009 \citep{NAS:2009} discussed the need for
determining error rates in methods proposed for firearms examination. As
seen in the case of most forensic applications, the first step for same
source matching and error rate determination involves identification of
unique features that are characteristic of the object at hand. For the
case of bullets and firearms, striation marks on the surface of the
bullet are considered to be such markings that can be used in methods
for same source matching. These marks are often a product of rifling and
impurities and defects due to manufacturing in the barrel of the gun,
which leads to engravings on the bullet surface
\citep{afte-article1992}. In current practice, firearm examiners
invariably make visual comparisons of bullet striae and use visual
assesment tools to dignify bullets as being matches and non-matches. One
way of accomplishing anykind of comparison between bullets is to do a
comparison between surface marking of two or more bullet lands. Bullet
Lands are considered to be areas between grooves made by the rifling
action of the barrel and the markings on them are considered to be
unique. The land engraved markings or sometimes termed as Bullet
profiles \citep{aoas} are striation marks made on Bullet lands and often
used for these land to land comparisons. Bullet Signatures is another
word used in literature as seen in the work of \citet{chu2013} and
\citet{aoas}. In our context bullet signatures refer to a processed
version of the raw land engraved markings or profiles. The generation of
bullet signatures involves first extraction of a bullet profile by
taking the cross-sectional of the surface at a given height and then
using loess fits to model the structure. The residuals of this fit are
called signatures, which are considered to be noise free and a good
reflection of the class charecteristics and unique features of a bullet.
A more detailed version of the extraction technique of signatures is
discussed by \citet{aoas}, where comprehensive details about the height
at which profile is to be selected, removing curvature, smoothing,
identifying groove locations are explained. Figure \ref{fig:rgl} shows
us how the extracted signature from a bullet land lines up with the
image of the land from which it was extracted. We can see also see in
the figure how the depth and relative position of the striation markings
seen in the image are interpreted as the signature.

@@

\begin{table}[!htb]
    \begin{minipage}{.5\linewidth}
      \caption{Chumbley et al. 2010}
      \centering 
\begin{tabular}{lrr}
\toprule
Classification & Match & Non-Match\\
\midrule
Match & 41 & 9\\
Non-Match & 2 & 48\\
\bottomrule
\end{tabular} \end{minipage}%
    \begin{minipage}{.5\linewidth}
      \centering
        \caption{Hadler et. al. (2017)} 
\begin{tabular}{lrr}
\toprule
Classification & Match & Non-Match\\
\midrule
Match & 47 & 3\\
Non-Match & 0 & 50\\
\bottomrule
\end{tabular} \end{minipage} 
\end{table}

>>>>

Same source analyses are a major part of an Forensic Toolmark Examiner's
job. In current practice examiners make these comparisons visually under
a comparison microscope and come to one of the following four
conclusions: identification, inconclusive, elimination or unsuitable for
examination \citet{afte-website-2}. These conclusions are made on the
basis of ``unique surface contours'' of the two toolmarks being in
``sufficient agreement'' \citep{afte-toolmarks1998}. AFTE describes the
term ``sufficient agreement'' as the possibility of another tool
producing the markings under comparison, as practically impossible
\citep{afte-toolmarks1998} \citep{afte-website}.
\%\gk{Although, they go on to concede that the interpretation of the identification depends on the examiner and is  therefore subjective in nature.}
This subjectivity in the assessment as well as the lack of error rates
are the main points of criticisms in the toolmark examination first
raised by the National Research Council in 2009 \citep{NAS:2009} and
later emphasized further by the President's Council of Advisors on
Science and Technology \citep{pcast2016}.

Technological advances, such as profilometers and confocal microscopy
allow to measure 3D surfaces \citep{vorburger2016} in a digitized form.
This forms the basis of statistical analysis of toolmarks. A statistical
approach based on data removes both subjectivity from the assessment and
allows a quantification of error rates for both false positive and false
negative identifications.

Various toolmarks have been studied in the literature: screwdriver marks
digitized using a profilometer have been analyzed in
\citet{manytoolmarks1} and \citet{chumbley}; \citet{manytoolmarks2} have
investigated 3D marks from screwdriver, tongue and groove pliers
captured using a confocal microscope; digitized marks from slip-joint
pliers generated by a surface profilometer have been investigated by
\citet{afte-chumbley}.

\citet{manytoolmarks2} define a relative distance metric and use it as
similarity measure between two toolmarks. \citet{manytoolmarks1} extract
many small segments in the markings of two toolmarks and compare
similiarity using a maximum pearson correlation coefficient. The
Chumbley scoring method, first introduced by \citet{chumbley}, uses a
similar but more extensive framework based on a Mann-Whitney U test of
the resulting correlation coefficients. This approach was later improved
by Hadler and Morris \citep{hadler}.
\hh{XXX describe the improvement in half a sentence XXX} In this paper,
we are investigating the applicability of the Chumbley scoring method by
\citet{hadler} to assess striation marks on bullet lands for same-source
identification.

Striation marks on bullets are made by impurities in the barrel. As the
bullet travels through the barrel, these imperfections leave
``scratches'' on the bullet surface. Typically, only striation marks in
the land engraved areas (LEAs) are considered \citet{afte-article1992}.
Bullet lands are depressed areas between the grooves made by the rifling
action of the barrel. Compared to toolmarks made by screwdrivers
striation marks on bullets are typically much smaller, both in length
and in width. Bullets also have a curved cross-sectional topography
\hh{sketch?}.

Bullet matching methods depend on the attribute being used for making
comparisons. For single feature based matching \citet{chu2013} used an
automatic method for counting consequtive matching striae (CMS). They
reported zero false positive rate but do not discuss about the false
negative rates for land to land matches. \citet{ma2004} and
\citet{vorburger2011} discuss about CCF (cross-correlation function) and
its discriminating power and application but do not talk about matches
and non-matches. \citet{aoas} use multiple attributes like CCF, CMS, D
(distance measure) etc in a random forrest based method and compare
every land against every other land of the Hamby dataset \citet{hamby}.
They report an out-of-bag error rate with false positive error rate of
0.0001 and false negative error rate of 0.2267 for 100 trees. For 300
trees they report correct prediction of all matches and non-matches.

The Chumbley score provides us with another approach in the same-source
assessment of bullet striation marks. \citet{chumbley}, in their paper,
compare two toolmarks with the intention of determining if it comes from
the same tool. The data for their study was obtained from 50
sequentially manufactured screwdriver tips.\citet{chumbley} reported
error rates for marking made by the tips at differnt angles. For a 30
degree they showed false negative error rate to be 0.023 and false
positive error rate be 0.09. For other angles the error rates remain the
same for false negatives and 0.01 for false positives. The improvement
proposed by \citet{hadler} on the other hand, did not consider effect of
angles and gave 0.06 false negative error rate and 0 false poisitive
error rate for markings made at the same angle.

\gk{\citep{afte-website} quotes sufficient agreement as "agreement of individual characteristics is of a quantity and quality that the likelihood another tool could have made the mark is so remote as to be considered a practical impossibility"}
\gk{could not find the pdf for AFTE theory of identification as it relates to toolmarks pdf. I believe it is not available for free. But from the AFTE Journal the abstract was available which mentions the term " sufficient agreement" as quoted above. Currently the website seems like the most reliable way of quoting AFTE, i have checked other sources in both occasions where the website has been quoated above, like an FBI paper available at the FBI website \citet{fbi-paper} and some non published but available pdf files in the AFTE website that have used these lines and quoted the AFTE journal. But since I cannot verify it in the Journal I have quoated the website.}
\gk{faden et al have not cited any paper that discusses the algorithm, i assume it was developed for the paper, the discussion of which is not as extensive as chumbley's. Its a variation of the chumbley method but from my understanding it is just the optimization step. and this paper came out much before the chumbley score.}

\hh{XXX first sentence: Same source analyses are a major part of an Forensic Toolmark examiners job. XXX second sentence: describe standard practice: In current practice firearm examiners  make these comparisons visually under a comparison microscope and report XXX what is being reported: AFTE. XXX find: AFTE theory of identification as it relates to toolmarks  }
\hh{One of the main criticisms in this toolmark examination first raised by the National Research Council in 2009 [@NAS:2009] and later emphasized further by the President's Council of Advisors on Science and Technology [@pcast2016] are the amount of subjectivity in the assessment and the lack of error rates. }
\hh{XXX we have digitized images, e.g. profilometers or Confocal microscopy allow us to measure the surface [@vorburger2016]}

@@

\hh{XXX In this paper, we are investigating the applicability of the Chumbley scoring method by @hadler to assessing striation marks on bullets for same-source identification. }

\hh{Striation marks on bullets are made by impurities in the barrel. As the bullet travels through the barrel, these imperfections leave "scratches" on the bullet surface. Typically, only striation marks in the land engraved areas (LEAs) are considered (XXX citation) [@afte-article1992]. XXX explain lands and grooves XXX. Compared to toolmarks made by screwdrivers striation marks on bullets are typically much smaller, both in length and in width. Bullets also have a curved cross-sectional topography. 
}

\hh{XXX How are bullets matched now? \citet{chu2013}: CMS, automatically - error rate?, using ccf XXX Vorburger paper XXX  and \citet{aoas}: multiple features, error rates }

\hh{The Chumbley score  provides us with another approach in the same-source assessment of bullet striation marks. }

\hh{end of intro: remaining paper is structured as follows: introduce to the data we get from confocal microscopy, introduce to profiles and signatures.
Introduce to chumbley score method, apply chumbley and discuss results ....
}

\subsection{Scans for land engraved
areas}\label{scans-for-land-engraved-areas}

\begin{itemize}
\tightlist
\item
  scans available: NIST database (citation), Hamby 44 and Hamby 252
  (Hamby citation)
\item
  move figure up
\item
  discuss cross section, profiles and signatures
\end{itemize}

Compairing pairs of toolmarks with the intention of matching it to a
tool has been studied many times in the past. Extensive examples can be
found in literature for tools and toolmark research ranging from
screwdrivers \citep{manytoolmarks1, manytoolmarks2, chumbley} to groove
pliers \citep{manytoolmarks2} to slip-joint pliers \citep{afte-chumbley}
and many more. In comparison to this, same source matching of bullets to
firearms has not been examined as prominently as that of toolmarks. Even
less information is available on validity of methods and error rates
associated with firearms examination. The National Academy of Sciences
in its report in 2009 \citep{NAS:2009} discussed the need for
determining error rates in methods proposed for firearms examination. As
seen in the case of most forensic applications, the first step for same
source matching and error rate determination involves identification of
unique features that are characteristic of the object at hand. For the
case of bullets and firearms, striation marks on the surface of the
bullet are considered to be such markings that can be used in methods
for same source matching. These marks are often a product of rifling and
impurities and defects due to manufacturing in the barrel of the gun,
which leads to engravings on the bullet surface
\citep{afte-article1992}. In current practice, firearm examiners
invariably make visual comparisons of bullet striae and use visual
assesment tools to dignify bullets as being matches and non-matches. One
way of accomplishing anykind of comparison between bullets is to do a
comparison between surface marking of two or more bullet lands. Bullet
Lands are considered to be areas between grooves made by the rifling
action of the barrel and the markings on them are considered to be
unique. The land engraved markings or sometimes termed as Bullet
profiles \citep{aoas,ma2004} are striation marks made on Bullet lands
and often used for these land to land comparisons. Bullet Signatures is
another word used in literature as seen in the work of \citet{chu2013}
and \citet{aoas}. In our context bullet signatures refer to a processed
version of the raw land engraved markings or profiles. The generation of
bullet signatures involves first extraction of a bullet profile by
taking the cross-sectional of the surface at a given height and then
using loess fits to model the structure. The residuals of this fit are
called signatures, which are considered to be noise free and a good
reflection of the class charecteristics and unique features of a bullet.
A more detailed version of the extraction technique of signatures is
discussed by \citet{aoas}, where comprehensive details about the height
at which profile is to be selected, removing curvature, smoothing,
identifying groove locations are explained. Figure \ref{fig:rgl} shows
us how the extracted signature from a bullet land lines up with the
image of the land from which it was extracted. We can see also see in
the figure how the depth and relative position of the striation markings
seen in the image are interpreted as the signature.

@@
